Trends in the contribution of major causes of death to the black-white life expectancy gap by US state

### 2017-08-03 Manuscript/Manuscript_COD_Trends_HealthandPlace.Rmd

---
title: Trends in the contribution of major causes of death to the black-white life expectancy gap by US state. A retrospective study of all deaths between 1969 and 2013
bibliography: bibliography.bib
output:
  word_document: default
  html_document: default
always_allow_html: yes
---

@@

**Objective:** To describe the pattern of contributions of six major causes of deaths to the gender-specific black-white life expectancy gap across US states over the last half-century, and identify divergent states.

@@

**Participants:** Using vital statistics and census data, we extracted the number of deaths and population sizes for the years 1969 to 2013, by state, gender, race, 19 age groups, and six major causes of death.

@@

**Results:** Although mortality from cardiovascular disease has decreased dramatically, its contribution to the life expectancy gap increased over time for men (from `r sprintf("%.1f", cvd.male.1969)` to `r cvd.male.2013` years), but decreased for women (from `r cvd.female.1969` to `r sprintf("%.1f", cvd.female.2013)` years). The contribution of non-communicable diseases to the gap was stable over time for men (approximately `r ncd.male.average` years) but decreased for women (from `r ncd.female.1969` to `r ncd.female.2013` years), while cancers exhibited an inverted-U trend for men (peaking at `r can.male.max` years in the early 1990s) and a stable contribution for women (approximately `r can.female.average` years). Both genders exhibited a decreased contribution from injuries (men: `r inj.male.1969` to `r inj.male.2013` years), that became negative for women (women: `r inj.female.1969` to `r inj.female.2013` years). Several states diverged from these general trends.

**Conclusions:** Life expectancy for blacks and whites has improved substantially in the US. For men, much of this improvement was due to narrowing differences in injury-related mortality, but these contributions were rivalled by an increasing gap in CVD-related mortality. In women, a crossover in injury-related mortality led to a narrower gap, realized partially by increasing mortality among whites.

###Strengths and limitations of this study

- The estimated trends for each state are relatively stable compared to estimates from smaller units of analysis such as counties, and can be used to study the effects of state-level policy in subsequent work.

- Examination of historical trends places the most recent differences in context, to disentangle whether the smallest state- and cause-specific contributions were historically small or underwent greater shifts over time compared to other states.

- The outcome of life expectancy integrates mortality rates across the entire age distribution making it sensitive to changes in mortality at any age.

Need to remove one of these:

- Due to data suppression when death counts were small, we did not consider more specific causes of death such as homicide and drug overdose.

- Migration between states and immigration to the US both affect state-level life expectancy and the black-white difference but were unmeasured.

- We examined the concept of race (black or white), not ethnicity (i.e., Hispanic), since ethnicity was unmeasured in the data for the earliest years and both poorly and differentially measured by state for many years thereafter.

@@

Life expectancy is a barometer for overall population health. In 2014, life expectancy at birth in the United States was 78.8 years, but hidden within this summary are important subgroups differences by self-defined racial ethnicity. Whites lived 3.4 years longer than blacks.^1^ While both black and white Americans have experienced increases in life expectancy since the beginning of the 20^th^ century, the difference between their life expectancies has varied.^2^

During the 1980s, the life expectancy gap widened and then narrowed in the 1990s due largely to relative mortality improvements among blacks in deaths due to homicide, HIV/AIDS, unintentional injuries, and, among women, due to differential improvements in heart disease.^3^ The gap continued to narrow during the 2000s, but states differed markedly in the magnitude of improvement. For example, the life expectancy gap in New York decreased from 8.0 to 2.4 years between 1990 and 2009, but in California the decrease was much more modest -- from 6.7 to 5.6 years.^4^ However, the extent to which different causes of death may have contributed to state-level changes in the black-white gap are unknown.

The over-arching objective of this study is to describe the contribution of selected causes of death to the black-white life expectancy gap over forty-five years, and identify divergent states and gender differences. To do so, we estimated state-level trends in the contribution of six major causes of death to the black-white life expectancy gap for men and women. This facilitates the identification of common reasons for the decreasing life expectancy gap and the identification of anomalous states. This study may therefore facilitate the identification of policies or social conditions in highly successful states, and suggest broad targets for interventions to further reduce the gap in less successful states.

@@

The data structure produced two statistical challenges that were addressed by our modeling strategy. Firstly, because of confidentiality, NCHS suppresses counts of deaths between 1 and 9,^9^ that we needed to impute. Secondly, several states had small black populations which produced sparse counts in some state-year-race-sex-age-cause strata. In these cases, the estimation of the mortality rates is imprecise, and can produce abrupt changes in the estimated mortality rates over time that are unrealistic. To better estimate the underlying mortality rates, we used a method that produces smoother age-specific mortality trends.

@@

To calculate life expectancy at birth, the smoothed cause-specific death counts were aggregated to estimate the total deaths in each age group, and standard life table methods were applied.^10,11^ Life expectancy at birth (in years) was calculated for blacks and whites in each state and year, by gender. The difference in life expectancy between blacks and whites was calculated by subtracting the black life expectancy from the white life expectancy in each state, year, and gender.

@@

To describe the national trend in each causes' contribution to the life expectancy gap over time, we used random effects meta regression to model each state's mean contribution estimated from the previous step as a function of a state random effect and a cubic spline for year (estimated using four internal knots). The model was inversely-weighted by the standard error of the mean estimate in each state-year. To identify states divergent from the national trend, we compared the state-specific trends estimated in the previous step to the national pattern estimated by setting the random effect to zero. These estimates relied on the R *lme4* package for random effects modeling, and the *splines* package to define cubic splines.

@@

Cardiovascular disease (CVD) is the most common cause of death for both men and women of both races. Mortality from CVD decreased markedly since 1969 for all groups (**Figures S3 and S4**), but white men underwent relatively larger mortality decreases than black men. As a result, the contribution of CVD to the black-white life expectancy gap among men increased modestly for most states from `r sprintf("%.1f", cvd.male.1969)` to `r cvd.male.2013` years (**Figure 1A**). Black women experienced relatively larger mortality decreases over time compared to white women, resulting in a decrease to the CVD contribution to the life expectancy gap from `r cvd.female.1969` to `r sprintf("%.1f", cvd.female.2013)` years nationally (**Figure 1B**). For both genders, Michigan, Wisconsin, California, and DC demonstrated higher-than-average increases in the contribution of CVD over time, whereas Massachusetts and New York showed persistently lower CVD contributions.

@@

Trends in mortalilty from injuries differ for blacks and whites. Over time, injury-related mortality has decreased precipitously among blacks, although an epidemic of homicide resulted in heightened mortality from injuries among black men during the 1980s and early 1990s (**Figure S9**).^3^ For whites, most states exhibit an initial period of gradual decreases in injury mortality, followed by a period of gradual increases, especially among white women (**Figure S10**). Thus, injury mortality rates among blacks are equivalent, or lower than those among whites in many states, with white women demonstrating sharper recent increases. The contribution of injuries to the mortality gap has therefore decreased over time from `r inj.male.1969` to `r inj.male.2013` years among men. Among women, the contribution went from `r inj.female.1969` in 1969 to `r inj.female.2013` in 2013, becoming negative in those states that now have lower injury-related mortality among blacks than whites (**Figures 4A and 4B**). Among men, DC, Illinois, Michigan and Missouri exhibit the largest recent injury-related contributions. Among women, many southern states exhibit negative contributions (indicating lower injury-related mortality among black women), while DC, Illinois, and Michigan having higher contributions than the general trend.

@@

**All other causes**

@@

We examined state-level trends in six major cause-of-death contributions to the black-white life expectancy gap over a 45-year period. While we found that many states conformed to the national pattern of cause-of-death specific contributions, several states diverged in ways that either resulted in the near or total elimination of the life expectancy gap, or alternatively its exacerbation. Overall, the racial gap in life expectancy is now smaller in many states in the South and Southeast that are associated historically with greater degrees of inequality and slower social progress, compared to the Midwest, which had been the receiving point of blacks from the South during “the Great Migration” for improved economic opportunities between 1916 and 1970.^14^ Trends in the black-white life expectancy gap among women in “rust-belt” states such as Michigan, Wisconsin and Illinois appear especially discouraging, compared to remarkable progress in Florida, North Carolina, Georgia and Mississippi. On the one hand, it appears to run counter to the narrative of worse overall health progress in the former Confederacy.^15^ On the other hand, greater racial equality can be achieved by falling black mortality rates or rising white rates, and so the closing gap cannot be celebrated without a careful examination of the two individual trends.^16^

Consistent with what has been reported at the national level,^17^ we found that CVD makes the largest contribution to the life expectancy gap, even though both races and genders experienced decreased CVD mortality rates over time. Injuries (especially homicide) continue to play a role in contributing to the current gap for men in several states, especially in the Midwest and DC, while contributing to a decreased gap among women in the South. This decreased gap among women likely reflects the higher rate of mortality from drug overdoses among whites compared to blacks during the ongoing opioid epidemic.^18^ Communicable disease continues to contribute to the gap in most states, especially in DC, New Jersey, and Florida. In our previous work, approximately half of the contribution of communicable disease to the gap was found to be due to HIV in 2008 nationally,^17^ suggesting that these states may increase equality through interventions targeted at reducing mortality from HIV. Cancer is also responsible for large contributions to the life expectancy gap in some states (Illinnois, Michigan, and Wisconsin) but not in others (New York, Massachusetts). Overall, New York and Massachusetts have virtually eliminated the black-white life expectancy gap for both men and women. However, both states began with only a minimal contribution of CVD and cancer to the mortality gap. The difference in early contributions is especially stark for the contribution of CVD among women in 1969, where many states, especially those in the South, had CVD contributions of more than two years. Thus, the reason for New York’s and Massachusetts’ low CVD contributions does not appear to be the effect of recent changes, but highlights a long-standing pattern of greater equality and points to the importance of viewing the most recent estimates in light of historical patterns.

We used data over 45 years, across two racial groups, and for all ages. By examining a wide time window, trends were examined that provided context for the contributions estimated in the most recent year. This allowed us to recognize when the estimated contributions for a particular cause of death were relatively low because they started low at the beginning of the follow-up, versus being low because they had decreased over time. Furthermore, our analysis was centered on decomposing life expectancy at birth, a measure that integrates mortality rates across all ages. This is especially important in light of recent studies that have received criticism for narrowly focusing on a specific age group.^19,20^ The Arriaga decomposition method accounts for the fact that deaths occurring in one cause stratum are reflected in the absence of that death in another stratum. In contrast to some recent work focused on county-level inequalities,^21,15^ states have some substantial advantages as analysis units; They have large populations to generate stable estimates, tend to be more demographically stable over time, and they are policy-making entities, and so have divergent social agendas that are reflected in varying patterns of mortality by race.

Our analysis has three main limitations. Firstly, we did not account for Hispanic ethnicity, since ethnicity was not measured until 1978, and even then it was poorly measured on the death certificate in many states for many years and our previous sensitivity analysis including Hispanics did not yield a substantial difference in the results.^8^ Had Hispanics been excluded from this study, the life expectancy estimates for both blacks and whites are expected to be lower, since Hispanics are known to have higher life expectancies than blacks and whites.^22^ Secondly, our analysis does not account for immigration by the foreign-born, or migration of native-born individuals between states. US immigrants have been shown to have life expectancies at birth that were approximately 5 years longer in 2010.^23^ Thus, the black-white gap in a state may decrease as more immigrants identify as black. Migration between states can also affect the estimated life expectancy of a state if it is associated with underlying risks of death. Migration is particularly important for the District of Columbia, because of its small geographical bounds. Affluent black residents have been shown to move to the suburbs in Maryland and Virginia, while affluent whites are migrating to DC. Both changes would serve to increase the racial gap in DC.^24^ Thirdly, by using broad causes of death we are unable to investigate trends in more specific causes. For example, injury mortality includes homicide, suicide, and drug overdoses, which are known to vary by sex, race, and over time. Drug overdose as a more specific cause of death has played an increasingly important role in shaping disparities in many states, but this is not separately identifiable in our analyses.^25,26^

The most important implications for a state-level analysis are using the patterns to make connections between changes to social policy and health outcome changes over time. States pursue varying approaches to an array of relevant programs and laws including tobacco control,^27^ Medicare coverage,^28^ hand-gun violence,^29^ and nutrition.^30^ These presumably affect health and longevity, and can have especially important impacts on racial differences when the policies are targeted to disadvantaged groups or outcomes that differentially affect blacks.^31^ The analyses reported here are entirely descriptive, but should be used for subsequent work that seeks to identify the causal processes that link social and health policies to successful trajectories in some states, and worsening disparities in others. Variation in policies over time can be used in quasi-experimental designs to uncover evidence for the role of specific policy innovations.^32^

`Communicable.Female.plot`[[3]] +
  geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 0.5), lwd = 0.5, col = "grey", alpha = 0.5) + 
  geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 1.5), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey", alpha = 0.5)
```

@@

@@

**needs to be reordered**

>>>>

---
title: Trends in the contribution of major causes of death to the black-white life expectancy gap by US state
bibliography: bibliography.bib
output:
  word_document: default
  html_document: default
always_allow_html: yes
---

@@

**Objective:** To describe the pattern of contributions of six major causes of deaths to the black-white life expectancy gap across US states over the last half-century, and identify divergent states.

@@

**Participants:** Using vital statistics and census data, we extracted the number of deaths and population sizes for the years 1969 to 2013, by state, gender, race, 19 age groups, and six major cause-of-death groups.

@@

**Results:** Although mortality from cardiovascular disease has decreased dramatically, its contribution to the black-white life expectancy gap increased over time for men (from `r sprintf("%.1f", cvd.male.1969)` to `r cvd.male.2013` years), but decreased for women (from `r cvd.female.1969` to `r sprintf("%.1f", cvd.female.2013)` years). The contribution of non-communicable diseases to the gap was stable over time for men (approximately `r ncd.male.average` years) and decreased for women (from `r ncd.female.1969` to `r ncd.female.2013` years), while cancers exhibited an inverted-U trend for men (peaking at `r can.male.max` years in the early 1990s) and a stable contribution for women (approximately `r can.female.average` years). Both genders exhibited a decreased contribution from injuries (men: `r inj.male.1969` to `r inj.male.2013` years), that became negative for women (women: `r inj.female.1969` to `r inj.female.2013` years). Several states diverged from these general trends.

**Conclusions:** The black-white life expectancy gap has decreased substantially in the US since 1969. For men, much of this improvement was due to narrowing differences in injury-related mortality, but these contributions were rivalled by an increasing gap in CVD-related mortality. In women, a crossover in injury-related mortality led to a narrower gap, due partially to increasing mortality among whites.

@@

Life expectancy is a barometer for overall population health. In 2014, life expectancy at birth in the United States was 78.8 years, but hidden within this summary are important subgroups differences. Whites lived 3.4 years longer than blacks.^1^ While both black and white Americans have experienced increases in life expectancy since the beginning of the 20^th^ century, the difference between their life expectancies has varied over time.^2^

During the 1980s the life expectancy gap widened, but then narrowed in the 1990s due largely to relative mortality improvements among blacks in deaths from homicide, HIV/AIDS, unintentional injuries, and, among women, from differential improvements in heart disease.^3^ The gap continued to narrow during the 2000s, but US states differed markedly in the magnitude of improvement. For example, the life expectancy gap in New York decreased from 8.0 to 2.4 years between 1990 and 2009, but in California the decrease was much more modest -- from 6.7 to 5.6 years.^4^ However, the extent to which different causes of death may have contributed to state-level changes in the black-white gap are unknown.

The overarching objective of this study is to describe the contribution of selected causes of death to the black-white life expectancy gap over forty-five years, and identify divergent states and gender differences. To do so, we estimated state-level trends in the contribution of six major causes of death to the black-white life expectancy gap for men and women. This facilitates the identification of common reasons for the decreasing life expectancy gap and the identification of anomalous states. This study may therefore facilitate the identification of policies or social conditions in highly successful states, and suggest broad targets for interventions to further reduce the gap in less successful states.

@@

The data structure produced two statistical challenges that were addressed by our modeling strategy. Firstly, because of confidentiality, NCHS suppresses counts of deaths between 1 and 9,^9^ which we needed to impute. Secondly, several states had small black populations which produced sparse counts in some state-year-race-sex-age-cause strata. In these cases, the estimation of the mortality rates is imprecise, and can produce abrupt changes in the estimated mortality rates over time that are unrealistic. To better estimate the underlying mortality rates, we used a method that produces smoother age-specific mortality trends.

@@

To calculate life expectancy at birth, the smoothed cause-specific death counts were aggregated to estimate the total number of deaths in each age group, and standard life table methods were applied.^10,11^ Life expectancy at birth (in years) was calculated for blacks and whites in each state and year, by gender. The difference in life expectancy between blacks and whites was calculated by subtracting the black life expectancy from the white life expectancy in each state, year, and gender.

@@

To describe the national trend in each cause's contribution to the life expectancy gap over time, we used random effects meta regression to model each state's mean contribution estimated from the previous step as a function of a state random effect and a cubic spline for year (estimated using four internal knots). The model was inversely-weighted by the standard error of the mean estimate in each state-year. To identify states divergent from the national trend, we compared the state-specific trends estimated in the previous step to the national pattern estimated by setting the random effect to zero. These estimates relied on the R *lme4* package for random effects modeling, and the *splines* package to define cubic splines.

A replication data set including the raw data and statistical code to reproduce the analysis and manuscript is publicly available on Github: link [DOI: removed for double blind review]

@@

Cardiovascular disease (CVD) is the most common cause of death for both men and women of both races. Mortality from CVD decreased markedly since 1969 for all groups (**Figures S3 and S4**), but white men underwent relatively larger mortality decreases than black men. As a result, the contribution of CVD to the black-white life expectancy gap among men increased modestly for most states from `r sprintf("%.1f", cvd.male.1969)` to `r cvd.male.2013` years (**Figure 1A**). Black women experienced relatively larger CVD mortality decreases over time compared to white women, resulting in a decrease of CVD to the contribution to the life expectancy gap from `r cvd.female.1969` to `r sprintf("%.1f", cvd.female.2013)` years nationally (**Figure 1B**). For both genders, Michigan, Wisconsin, California, and DC demonstrated higher-than-average increases in the contribution of CVD over time, whereas Massachusetts and New York showed persistently lower CVD contributions.

@@

Trends in mortalilty from injuries differ for blacks and whites. Over time, injury-related mortality has decreased precipitously among blacks, although an epidemic of homicide resulted in heightened mortality from injuries among black men during the 1980s and early 1990s (**Figure S9**).^3^ For whites, most states exhibit an initial period of gradual decreases in injury mortality, followed by a period of gradual increases, especially among white women (**Figure S10**). Thus, injury mortality rates among blacks are equivalent to, or lower than those among whites in many states, with white women demonstrating sharper recent increases. The contribution of injuries to the mortality gap has therefore decreased substantially over time from `r inj.male.1969` to `r inj.male.2013` years among men. Among women, the contribution went from `r inj.female.1969` in 1969 to `r inj.female.2013` in 2013, becoming negative in those states that now have lower injury-related mortality among blacks than whites (**Figures 4A and 4B**). Among men, DC, Illinois, Michigan and Missouri exhibit the largest recent injury-related contributions. Among women, many southern states exhibit negative contributions (indicating lower injury-related mortality among black women), while DC, Illinois, and Michigan having higher contributions than the general trend.

@@

`Communicable.Female.plot`[[3]] +
  geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 0.5), lwd = 0.5, col = "grey", alpha = 0.5) + 
  geom_hline(aes(yintercept = 1), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 1.5), lwd = 0.5, col = "grey", alpha = 0.5) +
  geom_hline(aes(yintercept = 2), lwd = 0.5, col = "grey", alpha = 0.5)
```
**All other causes**

@@

@@

We examined state-level trends in six major cause-of-death contributions to the black-white life expectancy gap over a 45-year period. While we found that many states conformed to a general pattern, several states diverged in ways that either resulted in the near or total elimination of the life expectancy gap, or alternatively its exacerbation. Overall, the racial gap in life expectancy is now smaller in many states in the South and Southeast that are associated historically with greater degrees of inequality and slower social progress, compared to the Midwest, which had been the receiving point of blacks from the South during “the Great Migration” for improved economic opportunities between 1916 and 1970.^14^ Trends in the black-white life expectancy gap among women in “rust-belt” states such as Michigan, Wisconsin and Illinois appear especially discouraging, compared to remarkable progress in Florida, North Carolina, Georgia and Mississippi. On the one hand, it appears to run counter to the narrative of worse overall health progress in the former Confederacy.^15^ On the other hand, because greater racial equality can be achieved by either falling black mortality rates or rising white rates, the reduced life expectancy gap cannot be celebrated without a careful examination of the two individual trends.^16^

Consistent with what has been reported at the national level,^17^ we found that CVD makes the largest contribution to the life expectancy gap, even though both races and genders experienced decreased CVD mortality rates over time. Injuries continue to play a role in contributing to the current gap for men in several states, especially in the Midwest and DC, while contributing to a decreased gap among women in the South. This decreased gap among women likely reflects the higher rate of mortality from drug overdoses among whites compared to blacks during the ongoing opioid epidemic.^18^ Communicable disease continues to contribute to the gap in most states, especially in DC, New Jersey, and Florida. In our previous work, approximately half of the contribution of communicable disease to the gap was found to be due to HIV in 2008 nationally,^17^ suggesting that these states may increase equality through interventions targeted at reducing mortality from HIV. Cancer is also responsible for large contributions to the life expectancy gap in some states (Illinnois, Michigan, and Wisconsin) but not in others (New York, Massachusetts). Overall, New York and Massachusetts have virtually eliminated the black-white life expectancy gap for both men and women. However, both states began with only a minimal contribution of CVD and cancer to the mortality gap. The difference in early contributions is especially stark for the contribution of CVD among women in 1969, where many states, especially those in the South, had CVD contributions of more than two years. Thus, the reason for New York’s and Massachusetts’ low CVD contributions does not appear to be the effect of recent changes, but highlights a long-standing pattern of greater equality and points to the importance of viewing the most recent estimates in light of historical patterns.

We used data over 45 years, across two racial groups, and for all ages. By examining a wide time window, trends were examined that provided context for the contributions estimated in the most recent year. This allowed us to recognize when the estimated contributions for a particular cause of death were relatively low because they started low at the beginning of the follow-up, versus being low because they had decreased over time. Furthermore, our analysis was centered on decomposing life expectancy at birth, a measure that integrates mortality rates across all ages. This is especially important in light of recent studies that have received criticism for narrowly focusing on a specific age group.^19,20^ The Arriaga decomposition method accounts for the fact that deaths occurring in one cause stratum are reflected in the absence of that death in another stratum. In contrast to some recent work focused on county-level inequalities,^21,15^ states have some substantial advantages as analysis units; They have large populations to generate stable estimates, tend to be more demographically stable over time, and they are policy-making entities with divergent social environments that are reflected in varying patterns of mortality by race.

Our analysis has three main limitations. Firstly, we did not account for Hispanic ethnicity, since ethnicity was not measured until 1978, and even then it was poorly measured on the death certificate in many states for many years. Our previous sensitivity analysis including Hispanics did not yield a substantial difference in the results for the overall life expectancy gap.^8^ Had Hispanics been excluded from this study, the life expectancy estimates for both blacks and whites are expected to be lower, since Hispanics are known to have lower mortality than blacks and whites.^22^ Secondly, our analysis does not account for immigration by the foreign-born, or migration of native-born individuals between states. US immigrants have been shown to have life expectancies at birth that were approximately 5 years longer in 2010.^23^ Thus, the black-white gap in a state may decrease as more immigrants identify as black. Migration between states can also affect the estimated life expectancy of a state if it is associated with underlying risks of death. Migration is particularly important for the District of Columbia, because of its small geographical bounds. Affluent black residents have been shown to move to the suburbs in Maryland and Virginia, while affluent whites are migrating to DC. Both changes would serve to increase the racial gap in DC.^24^ Thirdly, by using broad causes of death we are unable to investigate trends in more specific causes. For example, injury mortality includes homicide, suicide, and drug overdoses, which are known to vary by sex, race, and over time. Drug overdose as a more specific cause of death has played an increasingly important role in shaping disparities in many states, but this is not separately identifiable in our analyses.^25,26^

The most important implications for a state-level analysis are using the patterns to make connections between changes to social policy and health outcome changes over time. States pursue varying approaches to an array of relevant programs and laws including tobacco control,^27^ Medicare coverage,^28^ hand-gun violence,^29^ and nutrition.^30^ These presumably affect health and longevity, and can have especially important impacts on racial differences when the policies are targeted to disadvantaged groups or outcomes that differentially affect blacks.^31^ The analyses reported here are entirely descriptive, but could be used for subsequent work that seeks to identify the causal processes that link social and health policies to successful trajectories in some states, and worsening disparities in others. Variation in policies over time can be used in quasi-experimental designs to uncover evidence for the role of specific policy innovations.^32^

@@
